Visualizing Contingency Tables

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
AZ comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@519 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}).

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}
%FIXME% For the JSS paper there was criticism about haven MANET and Mondrian in the
%FIXME% same column, how did we decide ro resolve this? Different columns, I suppose...?!

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.
%FIXME% causality is not needed, hence omitted: in (hypothesized) causal relationships.

@@

%FIXME% Maybe formulate the following theory a bit more like:
%FIXME% "To fix notations, ...", "It is well-known that..."

When we try to explain data, we assume the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common hypotheses \cite{vcd:Agresti:2002} are

@@

\noindent It is easy to compute the \emph{expected} table under either
of these hypotheses. Consider a 
2-way contingency table with $I$ rows and $J$ columns,
cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
convenience, the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If the variables can be
separated into explanatory and dependent variable(s), the explanatory variable(s)
should be be used first for splitting to display the conditional distribution of
the dependent variable(s) given the explanatory variable(s). %FIXME% causality removed again
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary interest.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
\label{sec:colors} 
%FIXME% Maybe extend the footnote and say that the printed version is monochrome,
%FIXME% and that the electronic version uses color.
%FIXME% We can already say earlier that the R code for reproducing the figures
%FIXME% is available from a Web page. And then we can say here that the colored versions
%FIXME% can also be obtained from the Web page.
%FIXME% Where do we want to set up the URL? I would suggest
%FIXME%   http://statmath.wu-wien.ac.at/projects/vcd/

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improved}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}

\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A possible way of representing such a three-way table is 
to draw a mosaic plot of gender and
department and to highlight admission in the tiles (see Fig.~\ref{fig:ucbhighlight}).
From the widths of the tiles it can be seen that an
unequal number of students apply for the six
departments, and from the highlighting we see that the admission rates differ as
well among the departments. The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted. This result
is even better depicted by a doubledecker plot of admission, explained
by gender given department (see Fig.~\ref{fig:doubleucb}): here, the
difference of admission rates in department A is more marked.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbhighlight,fig=TRUE,echo=FALSE,height=6,width=9,results=hide>>=
mosaic(aperm(UCBAdmissions), gp = gpar(fill = grey.colors(2)),
   spacing = spacing_highlighting, split = TRUE, keep = FALSE, rep = FALSE)
@
\caption{Mosaic plot with highlighting for the \data{UCB admissions} data.}
\label{fig:ucbhighlight}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
%FIXME% right margin labels get clipped
<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(UCBAdmissions), col = rev(grey.colors(2)))
@
\caption{Dobledecker plot for the \data{UCB admissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

In our example, from the explorative analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figs.~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by using the conditioning variables first for splitting.
However, using a single plot, the strata are not corrected for the
marginal distribution of the stratifying variables. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to use trellis-like layouts for visualizing \emph{partial}
tables for given strata, defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding mosaics in the panels have the same size. 
In addition, trellis layouts help reducing the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} 
visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. 
% The overall deviation pattern
% thus indicates that in contrast to the impression given by the aggregated table, male
% students were not advantaged, and that in department A,
% \emph{female} students were even more successful than their male colleagues.

@@

To get an overview, we first plot a mosicplot of memory, age, and
education, and highlight attitude (see Fig.~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age. Most people in this group have only completed elementay
school. For people over 25, a majority has completed secondary
school. Among the younger people, fewer have memories of having been
punished as a child than among the older ones. But sad enough, 
for people with elementary or secondary school education,
a higher proportion \emph{with} memories tends to
have a moderate attitude to punishment than those without.
However, for people under 40, the proportion with
no attitude to punishment increases with the educational level.

@@

So both age and education seem to matter here. Is there any influence
of memory on attitude to punishment apart from that? To answer this
question, we test the hypothesis of memory and attitude being
conditionally independent, given age and education, and visualize the
model using an extended mosaic plot (see Fig.~\ref{fig:punishmosaic}). 
We can see that the model fits rather good, except for people over 25
with elementary school education: under the null, more people should
say ``no'' to physical punishment of children. This pattern can again also
be visualized using a trellis layout, this time using partial mosaic
plots (see Fig.~\ref{fig:punishmosaic}).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
tab_test <- coindep_test(tab, margin = c(3, 4))
mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
       expected = ~ (memory + attitude) * education * age,
       gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
       set_labels = list(age = c("15-24","25-39","40+")))
@
\caption{Mosaic plot for the \data{punishment} data.}
\label{fig:punishmosaic}
\end{center}
\end{figure}

>>>>

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|c|}
    \hline
 & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET} & \pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$& $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)& ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$& $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$& $\times$\\
    Extensible Design        &     &        & $\times$ &       &  &\\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.

@@

When we try to explain data, we assume the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common and well-known hypotheses \cite{vcd:Agresti:2002} are

@@

\noindent It is easy to compute the \emph{expected} table under either
of these hypotheses. To fix notations, in the following we consider a 
2-way contingency table with $I$ rows and $J$ columns,
cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
convenience, the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If the variables can be
separated into explanatory and dependent variable(s), the explanatory variable(s)
should be be used first for splitting to display the conditional distribution of
the dependent variable(s) given the explanatory variable(s). 
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary interest.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and
  Color Palettes\footnote{The printed version of the article is
    monochrome, but the electronic version uses colors. The colored
    versions of the plots are available from the code web page 
    (\url{http://statmath.wu-wien.ac.at/projects/vcd/}).}}
\label{sec:colors}

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improved}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is 
to draw a doubledecker plot nesting gender in
department and highlighting admission in the tiles (see Fig.~\ref{fig:doubleucb}).
From the widths of the tiles it can be seen that an
unequal number of students apply for the six
departments, and from the highlighting we see that the admission rates differ as
well among the departments (roughly speaking, they decrease from A to F). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

% \setkeys{Gin}{width=0.7\textwidth}
% \begin{figure}
% \begin{center}
% <<ucbhighlight,fig=TRUE,echo=FALSE,height=6,width=9,results=hide>>=
% mosaic(aperm(UCBAdmissions), gp = gpar(fill = grey.colors(2)),
%    spacing = spacing_highlighting, split = TRUE, keep = FALSE, rep = FALSE)
% @
% \caption{Mosaic plot with highlighting for the \data{UCB admissions} data.}
% \label{fig:ucbhighlight}
% \end{center}
% \end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
%FIXME% right margin labels get clipped
<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

In our example, from the explorative analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figs.~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by using the conditioning variables first for splitting.
However, using a single plot, the strata are not corrected for the
marginal distribution of the stratifying variables. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to use trellis-like layouts for visualizing \emph{partial}
tables for given strata, defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help reducing the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} 
visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. 
% The overall deviation pattern
% thus indicates that in contrast to the impression given by the aggregated table, male
% students were not advantaged, and that in department A,
% \emph{female} students were even more successful than their male colleagues.

@@

We first create a mosaic plot to get an overview on the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used last for splitting. Furthermore, we might 
expect Age and Education to have stratifying influence on the results, so they should
be used first for splitting.
In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Fig.~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age. Most people in this group have only completed elementay
school. For people over 25, a majority has completed secondary
school. Among the younger people, fewer have memories of having been
punished as a child than among the older ones. But sad enough, 
for people with elementary or secondary school education,
a higher proportion \emph{with} memories tends to
have a moderate attitude to punishment than those without.
However, for people under 40, the proportion with
no attitude to punishment increases with the educational level.

@@

So both age and education seem to matter here. But is there really any influence
of memory on attitude to punishment apart from that? Or, rephrased as a hypothesis:
are memory and attitude conditionally independent, given age and education?
We can visualize the model by means of a trellis layout, this time using partial mosaic
plots (see Fig.~\ref{fig:punishcotab}) for the strata defined by all factorial 
combinations of the age and education levels. In each panel, we plot mosaic plots of 
memory and attitude, where again the explanatory variable (memory) is used first 
for splitting.
The shading tells us the model seems to be consistent with the data, 
except for people over 25
with elementary school education: under the null, more people should have had
no attitude to physical punishment of children than actually had.

% ## \setkeys{Gin}{width=0.7\textwidth}
% ## \begin{figure}
% ## \begin{center}
% ## <<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
% ## tab_test <- coindep_test(tab, margin = c(3, 4))
% ## mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
% ##        expected = ~ (memory + attitude) * education * age,
% ##        gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
% ##        rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
% ##        set_labels = list(age = c("15-24","25-39","40+")))
% ## 
% ## \caption{Mosaic plot for the \data{punishment} data.}
% ## \label{fig:punishmosaic}
% ## \end{center}
% ## \end{figure}

The advantage of the exploratory view 
is to visualize the joint distribution of all variables.
On the other hand, the visualization might be 
strongly influenced by the marginal distribution of education over age (in particular
    the large proportion of 40+ people with elementary education)
 which is not of interest in the conditional independence problem. The
 conditional mosaic plot suppresses the effect of all stratifying
 variables (and their interactions).
